Gaps in stellar streams as a result of globular cluster fly-bys

### 2024-12-10 paper/main.tex
grammer and consistency edits

  Beyond the global (visible and dark) mass distribution, as we anticipated earlier in this Introduction, streams can be used also to infer the granularity of the dark matter, that is the mass and density of the sub-halos populating our Galaxy. According to simulations by \citet{2008MNRAS.391.1685S}, $\Lambda$-CDM predicts that galaxies grow hierarchically, with dark matter clumps forming at a wide range of masses and sizes. These clumps, or \textit{sub-halos}, are predicted to follow a mass distribution with a power-law slope slightly shallower than -2.0. To date, the smallest observed dark matter halo was detected through gravitational lensing in an Einstein ring and had a mass of $10^8$ solar masses, as described by \citet{2012Natur.481..341V}. However, some models predict that dark matter clumps could exist down to the mass of Earth-like planets at least \citep[see][ and discussion in \citet{2021arXiv211101148A} ]{2005JCAP...08..003G}.

@@

    For the galactic potential, we used the second model from \citet{2017A&A...598A..66P}, a superposition of a thin disk, thick disk, and dark matter halo, with masses and scale lengths provided in Table~1 of \citet{2023A&A...673A..44F}. This model is time-independent throughout our simulations. This model satisfies a series of observational constraints such as local solar stellar density, the galactic rotation curve, etc, similarly to other galactic models such as \texttt{MWpotential2014} from \citet{2015ApJS..216...29B} and \texttt{McMillian2017} from \citet{2017MNRAS.465...76M}. However, To balance data volume and computation time, we use only one galactic potential, which should suffice. \citet{2021MNRAS.505.5978V} found that only a few outer globular clusters are strongly affected by different potential models, with kinematic uncertainties being the dominant factor in orbital solution differences. Similarly, \citet{2024MNRAS.528.5189G} generated a globular cluster mass-loss catalog using seven different potential models and found their debris distributions to be model-independent, similarly to \citet{2023A&A...673A..44F}. While the clusters' positions may depend on the model, we assert that interaction rates and stream formation are unaffected by the choice of galactic potential model.

@@

    \subsubsection*{ \texttt{Full} simulations}
      There is a primary methodological departure from \citet{2023A&A...673A..44F}.  In that paper, globular clusters evolved under the gravitational effect of the Galaxy only, while now we include also the effect of all the other Galactic globular clusters, by taking into account the direct N-body interactions among them. To quantify these interactions, all clusters are represented by Plummer spheres, each with mass and half-mass radius as reported in the Baumgardt catalog \citep{2021MNRAS.505.5957B}. In the following of this paper, we refer to these simulations, where the globular clusters interactions are taken into account as \texttt{full} simulations. \\
    For these simulations, we proceed into two steps:
      \begin{enumerate}
        \item First, starting from the Galactocentric positions and velocities of all 165 Galactic globular clusters, we integrate their orbits back in time for 5~Gyr under the influence of the Galaxy itself, and their mutual influence. In the backward integration, the equations of motion for the globular cluster system are thus: 
        \begin{equation}
          \ddot{\vec{r}}_i = -\nabla \Phi(R_i,z_i) + \left.\sum_{j\neq i}^{N_{GC}} \frac{Gm_j}{\left(|\vec{r}_j - \vec{r}_i|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j - \vec{r}_i\right),
        \end{equation}\label{eq:GCNBody} where $\vec{r}$ indicates the galactocentric position vector, the index $i$ indicates the globular cluster of interest with $R_i,z_i$ being its galactocentric positions in cylindrical coordinates; the index $j$ indicates the other globular clusters that are summed over. $N_{GC}$ is the total number of globular clusters which in this study is 165, $m_j$ is the mass of the j-th cluster in the sample, $b_j$ is its Plummer scale radius, and $\vec{r_j}$ is its galactocentric position. $\Phi$ represents the Galactic smooth potential \citep[][Model~II, in the present case]{2017A&A...598A..66P}. Note that the masses and sizes of the globular clusters are kept constant in these simulations and not allowed to vary with time. This implies that we do not take into account in these models the internal evolution of the globular clusters, nor their response to the Galactic tidal field.

        \item Once the positions and velocities of the entire globular cluster system, as to 5~Gyr ago, are found, we sample Palomar~5 with  100,000 particles which follow a Plummer distribution of mass and half-mass radius as in the Baumgardt catalog.  We then integrate the evolution of these particles forward in time to the present day, by taking into account that each particle feels the gravitational potential of the Galaxy, that of its host cluster, and that of all the other clusters in the Galaxy. Note that we do not take into account self-gravity among particles. The particles experience the gravitational field yet do not contribute to it. The equation of motion of a generic particle among the 100,000 which populate Pal~5 is thus: 
        \begin{equation}
          \ddot{\vec{r}}_p = -\nabla \Phi(R_p,z_p) + \left.\sum_{j}^{N_{GC}} \frac{Gm_j}{\left(|\vec{r}_j(t) - \vec{r}_p|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j(t)- \vec{r}_p\right),
        \end{equation} \label{eq:equation_of_motion_particle} where the index $p$ represents one of the 100,000 particles of interest, $\vec{r_p}$ being its position, and $j$ indexes over the globular clusters as in Eq.~\ref{eq:GCNBody}. Note in that Eq.~\ref{eq:equation_of_motion_particle} the positions of the globular clusters are time-dependent since they are being loaded during this step and not computed, unlike Eq.~\ref{eq:equation_of_motion_particle}.

@@

    To quantify the impact of globular cluster passages to the density of Pal~5 stream, we performed a second set of simulations, which we refer to as the \texttt{reference} simulations in this paper.  These \texttt{reference} simulations use the same 50 sets of initial conditions as of the \texttt{full} simulations, the same galactic potential but exclude mutual interactions between globular clusters. The approach adopted for this second set of simulations is thus equivalent to that adopted already in \citet{2023A&A...673A..44F}. In Eq.~\ref{eq:GCNBody} only the gradient of the Galactic potential  is considered. In Eq.~\ref{eq:equation_of_motion_particle}, of the second term on the right-side of the equation, only the interaction of Pal~5 particles with Pal~5 itself are considered. In other words, sum iterates over only one globular cluster, the host. All the interactions with the other clusters in the Galaxy are indeed omitted.

@@

    To our knowledge, this is the first study to investigate the gravitational interactions between all Galactic globular clusters. While we did not  observe significant changes to orbital properties of the clusters, when subject to their mutual interactions -- a result which is  expected given the small sizes and vast distances between the clusters -- we do see clear effects of the presence of their gravitational field in the properties of Pal~5 stream. A particularly clear example of this effect is shown in Fig.~\ref{fig:stream_on_sky} which we selected for its prominent gaps. Two of these gaps are clearly visible in galactic coordinates and become even more apparent when marginalizing over latitude to reconstruct the 1D density profile of the stream, as a function of longitude.

    Regarding the shape of the density distribution, the central peak corresponds to the still-intact globular cluster whose stars have not yet escaped. The density peaks in the stream are of the same order as the cluster itself. This is inconsistent with reality, the cluster should be much more populated than the stream's peak density. Of course, this discrepancy is a result of our modeling choices. We use the present day mass and radius for Palomar~5 for the whole duration of the simulation. A consequence of this is that the system is modeled as being less dense than it should have been. In turn, our simulations have a strong initial mass which adds to the amplitudes of the profile density peaks of Fig.~\ref{fig:stream_on_sky}. This inaccuracy is acceptable for the scope of this work. First is that Palomar~5's tails indeed have more mass than the cluster itself. \citet{2017ApJ...842..120I} reports that there could be three times as much mass in Palomar 5's tails as the cluster itself. Secondly, the exact form of the density distribution is less important than having a population of particles present that can probe a cluster flyby event.

    To make a more quantitative comparison between the \texttt{reference} and \texttt{full} simulations, we work in the tail coordinate system, in which the stream is aligned with the cluster's orbit. The definition of this coordinate system is based on the work of \citet{2004AJ....127.2753D} and is shown in Fig.~\ref{fig:TailCoordinates}. Briefly, in this system the $x_{\textrm{tail}}$ coordinate represents the position of a particle along the orbit relative to the globular cluster—whether it is ahead (positive $x_{\textrm{tail}}$) or behind (negative $x_{\textrm{tail}}$) the cluster;  the $y_{\textrm{tail}}$ coordinate measures the particle's distance within the orbital plane, where positive values indicate that the particle is farther from the Galactic center and negative values indicate that it is closer. Fig.~\ref{fig:decomposition} shows a comparison between one of the 50 realizations of Palomar~5 stream taking into account the gravitational interactions with all other globular clusters in the Galaxy (top panel) and omitting them (bottom panel). This comparison clearly shows the presence of two wide ($\sim$ 100~pc and $\sim$ 1~kpc) gaps in the leading tail and of a more subtle under-density at $x_{\textrm{tail}}\sim 5$~kpc (we refer the reader to Appendix~\ref{sec:gap_detection} for a detailed description of the under-densities and gaps detection method). To find out which globular clusters were responsible of creating these gaps, and at what time the close passages took place, we estimated the gravitational acceleration along the orbit of Pal~5 and represented it in the ($t$, $\tau$) space. $t$ is the simulation time and $\tau$ is indicates how long it will take for Pal~5 to reach, or how  long ago it passed, a given point on its orbit. The use of $\tau$ is advantageous because the growth of the stream is approximately linear in $\tau$. On the other hand, in physical space streams on eccentric orbits indeed grow in time yet the growth rate is modulated with periodic expansion and contraction depending on the orbital phase. Adopting this time space, and reporting the gravitational acceleration along Pal~5 orbit in this space, the identification of the globular clusters which produced the perturbation, and the time at which it occurred becomes straight forward. We refer the reader to Appendix~\ref{sec:Perturber_Identification} for an in-depth description of the procedure. In this way, we could identify that the clusters responsible for creating gaps in the simulation of Palomar~5 stream reported in Fig.\ref{fig:decomposition} are NGC~2808, NGC~7078 and NGC~104, and that their close passages occurred 200~Myr, -1.9 and -2.1~Gyr ago, respectively. To further probe that the three clusters above are responsible of producing the gaps observed in this simulation, we conducted additional experiments by including only the perturbation of each of these three clusters at a time, and neglecting the gravitational perturbations of all the other clusters in the Galaxy. The results of these simulations are reported in the middle panels of Fig.~\ref{fig:decomposition} and clearly shows that NGC~2808, NGC~7078 and NGC~104 are the clusters responsible of creating the under-dense regions observed in the leading tails of Palomar~5. It is worth noting that the times at which the passages of these clusters occurred, according to our analysis, are in agreement with the observed width of the corresponding gaps: the encounter with NGC~2808 being very recent (only 200~Myr ago) its induced gap is still very thin, because it takes time for a perturbation to grow into an extended gap, as it is the case for those induced by the passages of  NGC~7078 and NGC~104, which occurred in earlier times. It is also interesting to emphasize that gaps as thin as those generated by the passage of NGC~2808, 200~Myr ago, can be detected by working in the tail coordinate system and by making a comparative analysis (\texttt{full} versus \texttt{reference} simulations): they are so thin that they cannot be directly identified in galactic coordinates (see Fig.~\ref{fig:stream_on_sky}).\\

>>>>

  Beyond the global visible and dark mass distribution streams can be used also to infer the granularity of the dark matter, that is the mass and density of the sub-halos populating our Galaxy. According to simulations by \citet{2008MNRAS.391.1685S}, $\Lambda$~Cold Dark Matter ($\Lambda$~CDM) predicts that galaxies grow hierarchically, with dark matter clumps forming at a wide range of masses and sizes. These clumps, or \textit{sub-halos}, are predicted to follow a mass distribution with a power-law slope slightly shallower than -2.0. To date, the smallest observed dark matter halo was detected through gravitational lensing in an Einstein ring and had a mass of $10^8$ solar masses, as described by \citet{2012Natur.481..341V}. However, some models predict that dark matter clumps could exist down to the mass of Earth-like planets at least \citep[see][ and discussion in \citet{2021arXiv211101148A} ]{2005JCAP...08..003G}.

@@

    For the galactic potential, we used the second model from \citet{2017A&A...598A..66P}, a superposition of a thin disk, thick disk, and dark matter halo, with masses and scale lengths provided in Table~1 of \citet{2023A&A...673A..44F}. This model is time-independent throughout our simulations. This model satisfies a series of observational constraints such as local solar stellar density, the galactic rotation curve, etc, similarly to other galactic models such as \texttt{MWpotential2014} from \citet{2015ApJS..216...29B} and \texttt{McMillian2017} from \citet{2017MNRAS.465...76M}. However, To balance data volume and computation time, we use only one galactic potential, which should suffice. \citet{2021MNRAS.505.5978V} found that only a few outer globular clusters are strongly affected by different potential models. Generally, kinematic uncertainties are the dominant factor in orbital solution differences per cluster. Similarly, \citet{2024MNRAS.528.5189G} generated a globular cluster mass-loss catalog using seven different potential models and found their debris distributions to be rather model-independent, similarly to \citet{2023A&A...673A..44F}. While the clusters' positions may depend on the model, we assert that interaction rates and stream formation are unaffected by the choice of galactic potential model.

@@

    \subsubsection*{ \texttt{Full} simulations}
    There is a primary methodological departure from \citet{2023A&A...673A..44F}.  In that paper, globular clusters evolved under the gravitational effect of the Galaxy only, while now we include also the effect of all the other Galactic globular clusters, by taking into account the direct N-body interactions among them. To quantify these interactions, all clusters are represented by Plummer spheres, each with mass and half-mass radius as reported in the Baumgardt catalog \citep{2021MNRAS.505.5957B}. For the remainder of this paper, we refer to these simulations, where the globular clusters interactions are taken into account as \texttt{full} simulations. \\
    For these simulations, we proceed into two steps:
      \begin{enumerate}
        \item First, starting from the Galactocentric positions and velocities of all 165 Galactic globular clusters, we integrate their orbits back in time for 5~Gyr under the influence of the Galaxy itself and their mutual influence. In the backward integration, the system of equations of motion for the globular clusters are thus: 
        \begin{equation}
          \ddot{\vec{r}}_i = -\nabla \Phi(R_i,z_i) + \left.\sum_{j\neq i}^{N_{GC}} \frac{Gm_j}{\left(|\vec{r}_j - \vec{r}_i|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j - \vec{r}_i\right),
        \end{equation}\label{eq:GCNBody} where $\vec{r}$ indicates the galactocentric position vector, the index $i$ indicates the globular cluster of interest with $R_i,z_i$ being its galactocentric positions in cylindrical coordinates; the index $j$ indicates the other globular clusters that are summed over. $N_{GC}$ is the total number of globular clusters which in this study is 165, $m_j$ is the mass of the j-th cluster in the sample, $b_j$ is its Plummer scale radius, and $\vec{r_j}$ is its galactocentric position. $\Phi$ represents the Galactic smooth potential \citep[][Model~II, in the present case]{2017A&A...598A..66P}. Note that the masses and sizes of the globular clusters are kept constant in these simulations and not allowed to vary with time. This implies that we do not take into account in these models the internal evolution of the globular clusters, nor their response to the Galactic tidal field.

        \item Once the positions and velocities of the entire globular cluster system, as to 5~Gyr ago, are found, we sample Palomar~5 with  100,000 particles which follow a Plummer distribution of mass and half-mass radius as in the Baumgardt catalog.  We then integrate the evolution of these particles forward in time to the present day, by taking into account that each particle feels the gravitational potential of the Galaxy, that of its host cluster, and that of all the other clusters in the Galaxy. Note that we do not take into account self-gravity among particles. The particles experience the gravitational field yet do not contribute to it. The equation of motion of a generic particle among the 100,000 which populate Pal~5 is thus: 
        \begin{equation}
          \ddot{\vec{r}}_p = -\nabla \Phi(R_p,z_p) + \left.\sum_{j}^{N_{GC}} \frac{Gm_j}{\left(|\vec{r}_j(t) - \vec{r}_p|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j(t)- \vec{r}_p\right),
        \end{equation} \label{eq:equation_of_motion_particle} where the index $p$ represents one of the 100,000 particles of interest, $\vec{r_p}$ being its position, and $j$ indexes over the globular clusters as in Eq.~\ref{eq:GCNBody}. Note in that Eq.~\ref{eq:equation_of_motion_particle} the positions of the globular clusters are time-dependent since they are being loaded during this step and not computed, unlike Eq.~\ref{eq:GCNBody}.

@@

    To quantify the impact of globular cluster passages to the density of Pal~5 stream, we performed a second set of simulations, which we refer to as the \texttt{reference} simulations in this paper.  These \texttt{reference} simulations use the same 50 sets of initial conditions as of the \texttt{full} simulations, the same galactic potential but exclude mutual interactions between globular clusters. The approach adopted for this second set of simulations is thus equivalent to that adopted already in \citet{2023A&A...673A..44F}. In Eq.~\ref{eq:GCNBody} only the gradient of the Galactic potential  is considered. In Eq.~\ref{eq:equation_of_motion_particle}, of the second term on the right-side of the equation, only the influence of Pal~5's Plummer sphere on Pal~5's particles is considered. In other words, sum iterates over only one globular cluster, the host. All the interactions with the other clusters in the Galaxy are indeed omitted.

@@

      % To our knowledge, this is the first study to investigate the gravitational interactions between all Galactic globular clusters. While we did not  observe significant changes to orbital properties of the clusters, when subject to their mutual interactions -- a result which is  expected given the small sizes and vast distances between the clusters -- we do 
    % This is no longer true after Sergey showed me the papers from Ischenko et al for the globular cluster ineractions

    We see clear effects of the presence of their gravitational field in the properties of Pal~5 stream. A particularly clear example of this effect is shown in Fig.~\ref{fig:stream_on_sky} which we selected for its prominent gaps. Two of these gaps are clearly visible in galactic coordinates and become even more apparent when marginalizing over latitude to reconstruct the 1D density profile of the stream, as a function of longitude.

    Regarding the shape of the density distribution, the central peak corresponds to the still-intact globular cluster whose stars have not yet escaped. The density peaks in the stream are of the same order as the cluster itself. This is inconsistent with reality, the cluster's peak density should be higher than the stream's peak density. Of course, this discrepancy is a result of our modeling choices. We use the present day mass and radius for Palomar~5 for the whole duration of the simulation. Consequently the system is modeled as being less dense than it should have been. In turn, our simulations have a strong initial mass which adds to the amplitudes of the profile density peaks of Fig.~\ref{fig:stream_on_sky}. This inaccuracy is acceptable for the scope of this work. First is that Palomar~5's tails indeed have more mass than the cluster itself. \citet{2017ApJ...842..120I} reports that there could be three times as much mass in Palomar 5's tails as the cluster itself. Secondly, the exact form of the density distribution is less important than having a population of particles present that can probe a cluster flyby event.

    To make a more quantitative comparison between the \texttt{reference} and \texttt{full} simulations, we work in the tail coordinate system, in which the stream is aligned with the cluster's orbit. The definition of this coordinate system is based on the work of \citet{2004AJ....127.2753D} and is shown in Fig.~\ref{fig:TailCoordinates}. Briefly, in this system the $x_{\textrm{tail}}$ coordinate represents the position of a particle along the orbit relative to the globular cluster. Positive values of $x_{\textrm{tail}}$ are ahead of the cluster while negative $x_{\textrm{tail}}$ is behind the cluster. The $y_{\textrm{tail}}$ coordinate measures the particle's distance within the orbital plane, where positive values indicate that the particle is farther from the Galactic center and negative values indicate that it is closer. Fig.~\ref{fig:decomposition} shows a comparison between one of the 50 realizations of Palomar~5 stream taking into account the gravitational interactions with all other globular clusters in the Galaxy (top panel) and omitting them (bottom panel). This comparison clearly shows the presence of two wide ($\sim$ 100~pc and $\sim$ 1~kpc) gaps in the leading tail and of a more subtle under-density at $x_{\textrm{tail}}\sim 5$~kpc (we refer the reader to Appendix~\ref{sec:gap_detection} for a detailed description of the under-densities and gaps detection method).

    To find out which globular clusters were responsible of creating these gaps, and at what time the close passages took place, we estimated the gravitational acceleration along the orbit of Pal~5 and represented it in the ($t$, $\tau$) space. $t$ is the simulation time and $\tau$ is indicates how long it will take for Pal~5 to reach, or how  long ago it passed, a given point on its orbit. The use of $\tau$ is advantageous because the growth of the stream is approximately linear in $\tau$. On the other hand, in physical space streams on eccentric orbits indeed grow in time yet the growth rate is modulated with periodic expansion and contraction depending on the orbital phase (see the top panel of Fig.~5 from \citet{2016MNRAS.457.3817S} for an example). Adopting this time space, and reporting the gravitational acceleration along Palomar~5 orbit in this space, the identification of the globular clusters which produced the perturbation, and the time at which it occurred becomes straight forward. We refer the reader to Appendix~\ref{sec:Perturber_Identification} for an in-depth description of the procedure. In this way, we could identify that the clusters responsible for creating gaps in the simulation of Palomar~5 stream reported in Fig.\ref{fig:decomposition} are NGC~2808, NGC~7078 and NGC~104, and that their close passages occurred 200~Myr, -1.9 and -2.1~Gyr ago, respectively.

    To further probe that the three clusters above are responsible of producing the gaps observed in this simulation, we conducted additional experiments by including only the perturbation of each of these three clusters at a time, and neglecting the gravitational perturbations of all the other clusters in the Galaxy. The results of these simulations are reported in the middle panels of Fig.~\ref{fig:decomposition} and clearly shows that NGC~2808, NGC~7078 and NGC~104 are the clusters responsible for creating the under-dense regions observed in the leading tails of Palomar~5. It is worth noting that the times at which the passages of these clusters occurred, according to our analysis, are in agreement with the observed width of the corresponding gaps: the encounter with NGC~2808 being very recent (only 200~Myr ago) its induced gap is still very thin, because it takes time for a perturbation to grow into an extended gap, as it is the case for those induced by the passages of  NGC~7078 and NGC~104, which occurred in earlier times. It is also interesting to emphasize that gaps as thin as those generated by the passage of NGC~2808, 200~Myr ago, can be detected by working in the tail coordinate system and by making a comparative analysis (\texttt{full} versus \texttt{reference} simulations): they are so thin that they cannot be directly identified in galactic coordinates (see Fig.~\ref{fig:stream_on_sky}).\\
